Experience Report: Haskell in Computational Biology

### 2012-06-27 mrfy_experience_report.tex
added a little sentence
Incorporated reviewer comments. Still need to cut some space.

Proteins are cellular machinery. They interact with one another and with other 
molecules to carry out the functions of living cells: metabolism, regulation, 
signaling, and so on.
A~protein's function is determined by its structure, 
and its structure is determined by the sequence of amino acids that
form the protein.
The amino-acid sequence is ultimately determined by a sequence of
nucleic acids in DNA, which we call a gene.
Given a genetic sequence, biologists wish to know the cellular
function of the protein that gene codes for.
The best known method of discovering such function is
to find other proteins of 
similar structure, which likely share similar function.
Proteins that share structure and function are expected to be
descended from a common ancestor---in biological terms, \emph{homologous}---and
thus
the problem of identifying proteins similar to a \textit{query sequence} is called 
\textit{homology detection}.

@@

A~hidden Markov model carries probabilities on some states and on all
state transitions.
Both the probabilities and the states are determined by the alignment:
\begin{itemize}
\item
For each column~$j$ of the alignment, the hidden Markov model has a
\emph{match state}~$M_j$.
The match state contains a table $e_{M_j}(x)$ which gives the
 probability that a homologous protein has amino acid~$x$ in
 column~$j$.
This probability is called an \emph{emission} probability.
\item
For each column~$j$ of the alignment, the hidden Markov model has a
\emph{deletion state}~$D_j$.
The deletion state determines the probability that the query sequence
has \emph{no} amino acid  in column~$j$, i.e., that it has a gap.
\item 
For each column~$j$ of the alignment, the hidden Markov model has an
\emph{insertion state}~$I_j$.
The insertion state contains a table $e_{I_j}(x)$ which gives the
probability that a homologous protein has gained amino acid~$x$ by
insertion at column~$j$.
\end{itemize}
The hidden Markov model also has distinguished \emph{begin} and \emph{end} states.
We~represent a state using an algebraic data type:
\remark{we promised a better name for HMMState. Thoughts?}
\begin{smallverbatim}
data HMMState = Mat | Ins | Del | Beg | End
\end{smallverbatim}

@@

The log of a probability is just a number.
To distinguish a transition probability from other numbers, we wrap it in
a value constructor:
\begin{smallverbatim}
type LogProb = Double
data TProb   = TProb { logProbability :: LogProb }
\end{smallverbatim}
We group transition probabilities into 7-tuples.
Each tuple corresponds to a ``node'' in the hidden Markov model
(a~dashed oval in \figref{plan7}) and to a column of the original
alignment.
As~noted above, 
a match state may transition to any state in the successor node;
an insertion state may transition to itself or to its successor match
state;
and
a~deletion may transition to its successor match
state or deletion state.
\begin{smallverbatim}
data TProbs = 
     TProbs { m_m :: TProb, m_i :: TProb, m_d :: TProb
            , i_m :: TProb, i_i :: TProb
            , d_m :: TProb, d_d :: TProb }
\end{smallverbatim}
In~our code, we select from this 7-tuple using the \texttt{edge}
function:
\smallverbatiminput{edge}
In MRFy, the first two arguments to \texttt{edge} are known at compile
time, and we expect \texttt{edge} to be inlined.

@@

Given the \texttt{Scored} abstraction, we can easily implement
Viterbi's equations for a match state.
To~implement the equation at the top right of \figref{viterbi},
we~choose the best solution
from among the three predecessor states;
we~use~\texttt{/+/} to add transition and emission scores,
and we use \texttt{fmap} to add \texttt{Mat} to the list of states
encountered up to this point:\remark{we promised to characterize the missing cases}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\smallverbatiminput{vfix}
Functions \texttt{tProb} and \texttt{edge} combine to select the
appropriate transition score (one of the $a'$~values).
Function~\verb+vee''+ is the \emph{memoized} version of~\verb+vee'+,
so
the call to~\verb+vee''+ produces the same result as a recursive call
to~\verb+vee'+, but faster:
\smallverbatiminput{memo}
Functions \texttt{Memo.memo3} and \texttt{Memo.arrayRange} come from
Luke Palmer's
\path{Data.MemoCombinators} package.
The value
\texttt{numNodes} represents~$\alignwidth$,
and \texttt{seqlen} represents~$N$.

@@

Our implementations of hidden Markov models and Viterbi's algorithm
duplicate existing functionality; these same algorithms were
implemented in C++ for HMMER and SMURF.
This reimplementation is a cost of using Haskell, and one that
other researchers may have to pay as well.
However, the authors of SMURF and HMMER implemented these algorithms
anew, even though existing implementations existed in C++.
In~MRFy, we use these algorithms as building blocks to implement
new functionality.

@@

Why do we have \texttt{pt0} and \texttt{nextPt}
instead 
of a single function with type \mbox{\texttt{Rand r [Scored pt]}},
which could produce an infinite list?
Because after the first move
the starting point for the next move depends on whether the first move
was useful.\remark{I hate this ``explanation.'' ---NR}
That decision is made by the \texttt{utility} function,
which scrutinizes a move represented as follows:
\smallverbatiminput{move}
The decision about utility uses not only a source of randomness but
also the \emph{age} of the \texttt{younger} point which is
moved~to.\remark{This word ``age'' is \emph{entirely} bogus!}
For example, in~simulated annealing, as the search progresses,
the \texttt{utility} function becomes less likely to express a move
that results in a worse score.\remark{And I hate this explanation
too.}

\noindent
\vbox{
Using these pieces, function \texttt{everyGen} produces an infinite
sequence containing a mix of useful and useless moves:
\smallverbatiminput{everygen}
}
\noindent
Because both \texttt{nextPt} and \texttt{utility} consume randomness,
the computation is very sequential, but it does use a little bit of
laziness:
from its starting point, \texttt{everyPt} provides an infinite list of
randomly chosen successor points, then calls \texttt{agedUtility} to
tag each one with its age and utility.\remark{There's that bogus age
again.}
The Prelude function \texttt{span} finds the first useful successor,
but in case there \emph{is} no successor, the code must also produce
all the useless successors.
If~we do find a useful successor, we start searching anew from that
point, with a recursive call to \texttt{everyPt}.

The most telling line of code in \texttt{everyPt} is the last line of
the body, which shows that the result begins with a useful point,
is followed by a (possibly infinite, possibly empty) list of useless
points, and then continues recursively with another useful point.
When the computation returned by \texttt{everyPt} is evaluated against
a random-number generator, it returns an infinite list containing a
mix of useful and useless points.
This mix can be passed to a test function of type \texttt{SearchStop~pt}:\\
\vbox{
\smallverbatiminput{strat}
\smallverbatiminputx{stop}
\smallverbatiminputx{search}
}
\noindent
We have implemented two kinds of stop functions...

N.B. WE HAVE TWO KINDS OF NON-MODULARITY.  FIRST, TERMINATION HAS TO
LOOK AT \emph{ALL} STATES, EVEN THOUGH WE WOULD PREFER THAT IT LOOK
ONLY AT USEFUL STATES.  SECOND, AGE IS EVERYWHERE.  IN PARTICULAR, WE
CANNOT SEPARATE THE AGE OF THE CURRENT STATE FROM THE UTILITY
DECISION.

\remark{We need to explain why we don't just generate an infinite lazy
list of thingies.  And perhaps we can consider an infinite tree!}

The \texttt{SearchStrategy} interface provides ample scope for
experiments.\remark{\P\ needs rewriting}
Random-mutation hill climbing took fifty lines of code and one~day to
implement.
Simulated annealing reuses hill climbing's
 \texttt{nextPt}, \texttt{pt0}, and \texttt{quit} functions.
Simulated annealing's  new \texttt{accept} function took ten
lines of code and half an hour to implement.
(Hill climbing accepts a point only if it scores better than its
predecessor;
simulated annealing may accept a point that scores worse.)
Our genetic algorithm reuses the same
\texttt{pt0} and \texttt{quit} functions, and its \texttt{accept}
function took only a few lines of code.
But~the recombination of parent
placements in its \texttt{nextPt} function took forty lines of code
and a full day to implement.
\ifnotcutting
Our~current \texttt{quit} function simply runs for a fixed number of
generations, but its interface will enable us to try other criteria,
such as stopping once $k$~generations have passed without more than $E$
improvement in score. 
\fi

\subsection*{The old search algorithm}

We~pass the
search function a \texttt{Scorer}, which runs Viterbi's algorithm.
\smallverbatiminput{orig-scoredecl}
A~search heuristic is characterized by four functions, which are
bundled into a record called \texttt{SearchStrategy}.
\smallfuzzverbatiminput{2.3pt}{orig-strategy}
In this record,
\begin{itemize}
\item
Function \texttt{pt0} produces the first generation of placements.
\item
Function \texttt{nextGen} produces a new generation of placements from
an existing generation.
Because scoring a placement is expensive, 
and some placements may persist from the old generation to the new
generation, \texttt{nextGen} works with \emph{scored} placements, not
bare placements.
\item
A~new generation produced by \texttt{nextGen} may be no better
than the generation it replaces.
When it is first produced, the new generation is ``on trial.''
The \texttt{accept} function decides whether to keep a new generation
or to try \texttt{nextGen} again with a different seed.
The \texttt{accept} function considers a list of type
\texttt{[Scored~Age]}.
At~the head of this list is the score of the best
placement of the trial generation, together with the \texttt{Age} of the trial
generation, which is the total number of generations tried.
The remaining elements of the list hold the history of scores and ages for
previously \emph{accepted} generations.
(The list omits generations that didn't survive the \texttt{accept} trial.)
For convenience, the \texttt{accept} function also receives
the \texttt{Age} of the trial generation.
\item
Finally, the \texttt{quit} function decides when to stop the search.
\ifnotcutting
It~too considers the list of accepted generations and the age of the current
generation. 
\fi
\end{itemize}
This \texttt{SearchStrategy} interface somewhat resembles an object-oriented
style of programming.
However, it is \textit{easier} to create a SearchStrategy
in Haskell than using conventional object-oriented techniques, because
all the functions in a SearchStrategy are partial applications of
other functions.
If we were using conventional object-oriented
techniques, we would have to declare subclasses with instance
variables and would have to write code to initialize objects of those
subclasses.

\seclabel{search}%
We hope you find the search simple and straightforward:

\smallfuzzverbatiminput{11pt}{orig-search}

@@

MRFy calls \verb+vee'+ several
 times
for each placement in each search generation.
Calls on query sequences between beta strands are independent,
as are calls on different placements.
We~found them easy to parallelize
using
\texttt{Control.Parallel}:
we substituted
\texttt{parmap rseq} for \texttt{map}.
Even when there is just one placement per generation,
MRFy can keep six cores busy.
Given multiple placements per generation,
it~can use forty-eight cores.
However, we have found that MRFy performs best, at least on our
hardware, with between 12 and 16 cores. We have access to only
one 48-core system, and while MRFy will keep all 48 cores busy, it
performs better when given no more than 16 cores.\remark{give specific
numbers here}

MRFy's predecessor, SMURF, does not perform stochastic search,
and thus its overall run-time performance cannot be compared directly to MRFy.
However, both programs implement the Viterbi algorithm.
On an exceedingly large, contrived example, comprising 56,000 amino acids
(larger than any known protein) and 343 nodes, and ignoring all beta-strand
information, SMURF's Viterbi implementation completed in 0.497 seconds on a
2.6GHz Core i7.
On the same benchmark, MRFy's Viterbi implementation required 1 minute 57
seconds.
Despite the inferior runtime performance, MRFy still meets our performance
goals.
This is because the stochastic search algorithm relies on the Viterbi algorithm
applied to many, significantly smaller hidden Markov models than SMURF's
multidimensional programming must tackle.
MRFy computes an alignment for a 12-stranded ``beta sandwich'' structure in
generally less than one minute, while SMURF consumes over 16GB of memory
and does not terminate even after eight hours.
Had we implemented the SMURF algorithm in Haskell, we believe that performance
would be unacceptable.
However, for the algorithm underlying MRFy, we are satisfied with the runtime
performance of Haskell and GHC.

In~\secref{viterbi} above, the example case for the \verb+vee'+
function computes not just a score but also an optimal path.
MRFy's \texttt{search} function, in \secref{search}, uses only the
score.
Even though Haskell evaluation is lazy, the \verb+vee'+ 
still allocates thunks that could compute paths.
We~cloned \verb+vee'+ and modified it to compute
only a score, with no path, and this change improved 
run-time performance by nearly~50\%.
But when \texttt{search} finally decides on a placement, we need a
version of \verb+vee'+ that \emph{does} compute a path, so we can
extract the optimal parse for that placement.

@@

Debugging led also not to awkward code, but to awkward interactions on
our team.
The two junior members of the team wanted to apply the debugging
skills they had honed through years of imperative programming.
The senior partner kept nattering on about QuickCheck properties.
No~progress was made until the first draft of this paper was prepared.
At~that point, we were quickly about to write QuickCheck
specifications for several properties of the Viterbi algorithm:
\begin{itemize}
\item
The total number of \verb+Mat+ and \verb+Ins+ states returned equals
the number of amino acids in the query sequence.
\item
The total number of matches, insertions, and deletions returned equals the
number of nodes in the hidden Markov model,
\emph{provided} that a maximal, nonempty, contiguous subsequence
of~\verb+Ins+ states counts as a single node,
and \verb+Mat+ and \verb+Del+ states count as one node each.
\item
The sequence returned satisfies the Plan7 invariant, that is, states
\verb+Ins+ and \verb+Del+ are never adjacent.
\end{itemize}
\textbf{These properties need to be followed by more ambitious
properties},
such as proving that nearby sequences are less likely than the
sequence returned by \verb+vee'+.

@@

We~modified Matt to use information about sequences as well as structure.
The modification added 2,000 lines of code, and it calls
external sequence aligners that we did not write.
We~thought the modification would take three months, 
but it took most of a~year.
Matt uses such
data structures such as mutable oct-trees, vectors, and arrays.
It~uses clever pointer arithmetic.
The mutable data structures were difficult to 
repurpose, and the pointer arithmetic was \emph{too} clever: 
nearly every change resulted in new segfaults.

@@

We~can also make a preliminary comparison of performance.
SMURF can parse a simple protein structure, optimally, in under 10~seconds.
It~can parse some more complex structures in 2--3~minutes.
But when asked to parse the most complex structures, it runs out of
memory and has to be killed, even on a system with 128GB of~RAM.
By~contrast, MRFy can parse any of these structures, approximately,
in 30~seconds.\remark{reconcile this vague paragraph with performance section above}
This performance
meets our goals, and it means that MRFy can be used to scan whole genomes.

Looking beyond our research group to computational biology more
broadly, our experience with other software is better.
Little of it is written in functional languages, 
but much of the software shared by the community is excellent.
MRFy's training component was derived from that of~HMMER,
and
working with HMMER 
codebase was pleasant;
data structures and their
mutators are well documented. 
There is a 
BioHaskell library, part of which we use,
but it is not nearly as 
complete as BioPython or BioRuby, which are heavily used in the community.
We hope that tools for computational biology in
Haskell continue to mature.

@@

An~anonymous referee\remark{Just one referee?} spurred us to think
more deeply about laziness.
Josef Svenningsson suggested a correspondence between \mrfy's search
and stream fusion, which led us eventually to the \texttt{Utility} type.
We also thank Lenore Cowen, Kathleen Fisher, Ben Hescott, Brad
Larsen, and Nathan Ricci. 
\fi

>>>>

Proteins are cellular machinery. They interact with one another and with other 
molecules to carry out the functions of living cells: metabolism, regulation, 
signaling, and so on.
A~protein's function is determined by its structure, 
and its structure is determined by the sequence of amino acids that
form the protein.
The amino-acid sequence is ultimately determined by a sequence of
nucleic acids in DNA, which we call a gene.
Given a genetic sequence, biologists wish to know the cellular
function of the protein that gene codes for.
One of the best known methods of discovering such function is
to find other proteins of 
similar structure, which likely share similar function.
Proteins that share structure and function are expected to be
descended from a common ancestor---in biological terms, \emph{homologous}---and
thus
the problem of identifying proteins similar to a \textit{query sequence} is called 
\textit{homology detection}.

@@

A~hidden Markov model carries probabilities on some states and on all
state transitions.
Both the probabilities and the states are determined by the alignment:
\begin{itemize}
\item
For each column~$j$ of the alignment, the hidden Markov model has a
\emph{match state}~$M_j$.
The match state contains a table $e_{M_j}(x)$ which gives the
 probability that a homologous protein has amino acid~$x$ in
 column~$j$.
This probability is called an \emph{emission} probability.
\item
For each column~$j$ of the alignment, the hidden Markov model has a
\emph{deletion state}~$D_j$.
The deletion state determines the probability that the query sequence
has \emph{no} amino acid  in column~$j$, i.e., that it has a gap.
\item 
For each column~$j$ of the alignment, the hidden Markov model has an
\emph{insertion state}~$I_j$.
The insertion state contains a table $e_{I_j}(x)$ which gives the
probability that a homologous protein has gained amino acid~$x$ by
insertion at column~$j$.
\end{itemize}
The hidden Markov model also has distinguished \emph{begin} and \emph{end} states.
We~represent the state chosen for each node using an algebraic data type:
\remark{we promised a better name for HMMState. Thoughts?}
\begin{smallverbatim}
data HMMState = Mat | Ins | Del | Beg | End
\end{smallverbatim}

@@

The log of a probability is just a number, which we call a \texttt{Score}.
A transition probability connects two \texttt{HMMState}s, and so we wrap it in
a value constructor:
\begin{smallverbatim}
newtype Score = Score Double

data TProb =
     TProb { logProbability :: Score
           , fromState :: HMMState
           , toState :: HMMState
           }

\end{smallverbatim}
We group transition probabilities into 7-tuples.
Each tuple corresponds to a ``node'' in the hidden Markov model
(a~dashed oval in \figref{plan7}) and to a column of the original
alignment.
As~noted above, 
a match state may transition to any state in the successor node;
an insertion state may transition to itself or to its successor match
state;
and
a~deletion may transition to its successor match
state or deletion state.
\begin{smallverbatim}
data TProbs = 
     TProbs { m_m :: TProb, m_i :: TProb, m_d :: TProb
            , i_m :: TProb, i_i :: TProb
            , d_m :: TProb, d_d :: TProb }
\end{smallverbatim}
In~our code, we select from this 7-tuple using the \texttt{edge}
function:
\smallverbatiminput{edge}
In MRFy, the first two arguments to \texttt{edge} are known at compile
time, and we expect \texttt{edge} to be inlined.

@@

Given the \texttt{Scored} abstraction, we can easily implement
Viterbi's equations for a match state.
To~implement the equation at the top right of \figref{viterbi},
we~choose the best solution
from among the three predecessor states;
we~use~\texttt{/+/} to add transition and emission scores,
and we use \texttt{fmap} to add \texttt{Mat} to the list of states
encountered up to this point:
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\smallverbatiminput{vfix}
The other cases (\texttt{Ins} and \texttt{Del}) look similar.
Functions \texttt{tProb} and \texttt{edge} combine to select the
appropriate transition score (one of the $a'$~values).
Function~\verb+vee''+ is the \emph{memoized} version of~\verb+vee'+,
so
the call to~\verb+vee''+ produces the same result as a recursive call
to~\verb+vee'+, but faster:
\smallverbatiminput{memo}
Functions \texttt{Memo.memo3} and \texttt{Memo.arrayRange} come from
Luke Palmer's
\path{Data.MemoCombinators} package.
The value
\texttt{numNodes} represents~$\alignwidth$,
and \texttt{seqlen} represents~$N$.

@@

Our implementations of hidden Markov models and Viterbi's algorithm
duplicate existing functionality; these same algorithms were
implemented in C++ for HMMER and SMURF.
This reimplementation is a cost of using Haskell, and one that
other researchers may have to pay as well, as fewer computational biology
libraries exist for Haskell. For example, the National Center for
Biotechnology Information publishes its own library of algorithms
for computational biology.
However, the authors of SMURF and HMMER implemented these algorithms
anew, even though existing implementations existed in C++.
In~MRFy, we use these algorithms as building blocks to implement
new functionality.

@@

Why do we have \texttt{pt0} and \texttt{nextPt}
instead 
of a single function with type \mbox{\texttt{Rand r [Scored pt]}},
which could produce an infinite list?
Because after the first move
the starting point for the next move depends on whether the first move
was useful, whereas the first move is always useful.
That decision is made by the \texttt{utility} function,
which scrutinizes a move represented as follows:
\smallverbatiminput{move}
The decision about utility uses not only a source of randomness but
also the \emph{age} of the search, which corresponds to the
computational cost to reach a point in the search.
For example, in~simulated annealing, as the search progresses,
the \texttt{utility} function becomes less likely to express a move
that results in a worse score.

\noindent
\vbox{
Using these pieces, function \texttt{everyGen} produces an infinite
sequence containing a mix of useful and useless moves:
\smallverbatiminput{everygen}
}
\noindent
Because both \texttt{nextPt} and \texttt{utility} consume randomness,
the computation is very sequential, but it does use a little bit of
laziness:
from its starting point, \texttt{everyPt} provides an infinite list of
randomly chosen successor points, then calls \texttt{agedUtility} to
tag each one with the search's age and the point's utility.
The Prelude function \texttt{span} finds the first useful successor,
but in case there \emph{is} no successor, the code must also produce
all the useless successors.
If~we do find a useful successor, we start searching anew from that
point, with a recursive call to \texttt{everyPt}.

The most telling line of code in \texttt{everyPt} is the last line of
the body, which shows that the result begins with a useful point,
is followed by a (possibly infinite, possibly empty) list of useless
points, and then continues recursively with another useful point.
When the computation returned by \texttt{everyPt} is evaluated against
a random-number generator, it returns an infinite list containing a
mix of useful and useless points.
This mix can be passed to a test function of type \texttt{SearchStop~pt}:\\
\vbox{
\smallverbatiminput{strat}
\smallverbatiminputx{stop}
\smallverbatiminputx{search}
}
\noindent

We have two kinds of non-modularity.
First, termination has to
look at \emph{all} states, even though we would prefer that it look
only at useful states.
This is because otherwise we risk non-termination:
we are not guaranteed a useful move after any finite number of steps.
Second, age of the search shows up in multiple places.
In particular, we cannot separate the age of the search at the
current state from the utility decision.

The \texttt{SearchStrategy} interface provides ample scope for
experiments.\remark{\P\ needs rewriting}
Random-mutation hill climbing took fifty lines of code and one~day to
implement.
Simulated annealing reuses hill climbing's
 \texttt{nextPt}, \texttt{pt0}, and \texttt{quit} functions.
Simulated annealing's  new \texttt{accept} function took ten
lines of code and half an hour to implement.
(Hill climbing accepts a point if and only if it scores better than its
predecessor;
simulated annealing may accept a point that scores worse.)
Our genetic algorithm reuses the same
\texttt{pt0} and \texttt{quit} functions, and its \texttt{accept}
function took only a few lines of code.
But~the recombination of parent
placements in its \texttt{nextPt} function took forty lines of code
and a full day to implement.
\ifnotcutting
Our~current \texttt{quit} function simply runs for a fixed number of
generations, but its interface will enable us to try other criteria,
such as stopping once $k$~generations have passed without more than $E$
improvement in score. 
\fi

Our \texttt{SearchStrategy} interface somewhat resembles an object-oriented
style of programming.
However, it is \textit{easier} to create a SearchStrategy
in Haskell than using conventional object-oriented techniques, because
all the functions in a SearchStrategy are partial applications of
other functions.
If we were using conventional object-oriented
techniques, we would have to declare subclasses with instance
variables and would have to write code to initialize objects of those
subclasses.

@@

MRFy calls \verb+vee'+ several
 times
for each placement in each search generation.
Calls on query sequences between beta strands are independent,
as are calls on different placements.
We~found them easy to parallelize
using
\texttt{Control.Parallel}:
we substituted
\texttt{parmap rseq} for \texttt{map}.
Even when there is just one placement per generation,
MRFy can keep six cores busy.
Given multiple placements per generation,
it~can use forty-eight cores.
However, we have found that MRFy performs best, at least on our
hardware, with between 12 and 16 cores.
We have access to only
one 48-core system, and while MRFy will keep all 48 cores busy, it
performs only slightly better, at an average of 35 seconds to find an alignment
on a large protein structure, than when given 16 cores, which takes an average
of 37 seconds.

MRFy's predecessor, SMURF, does not perform stochastic search,
and thus its overall run-time performance cannot be compared directly to MRFy.
However, both programs implement the Viterbi algorithm.
On an exceedingly large, contrived example, comprising 56,000 amino acids
(larger than any known protein) and 343 nodes, and ignoring all beta-strand
information, SMURF's Viterbi implementation completed in 0.497 seconds on a
2.6GHz Core i7.
On the same benchmark, MRFy's Viterbi implementation required 1 minute 57
seconds.
Despite the inferior runtime performance, MRFy still meets our performance
goals.
This is because the stochastic search algorithm relies on the Viterbi algorithm
applied to many, significantly smaller hidden Markov models than SMURF's
multidimensional programming must tackle.
MRFy computes an alignment for a 12-stranded ``beta sandwich'' structure in
generally less than one minute, while SMURF consumes over 16GB of memory
and does not terminate even after eight hours.
Had we implemented the SMURF algorithm in Haskell, we believe that performance
would be unacceptable.
However, for the algorithm underlying MRFy, we are satisfied with the runtime
performance of Haskell and GHC.
Nonetheless, we expect that we can still improve MRFy's run-time performance
with further profiling and tuning.

In~\secref{viterbi} above, the example case for the \verb+vee'+
function computes not just a score but also an optimal path.
MRFy's \texttt{search} function, in \secref{mrfy}, uses only the
score.
Even though Haskell evaluation is lazy, the \verb+vee'+ 
still allocates thunks that could compute paths.
We~cloned \verb+vee'+ and modified it to compute
only a score, with no path, and this change improved 
run-time performance by nearly~50\%.
But when \texttt{search} finally decides on a placement, we need a
version of \verb+vee'+ that \emph{does} compute a path, so we can
extract the optimal parse for that placement.

@@

Debugging led also not to awkward code, but to awkward interactions on
our team.
The two junior members of the team wanted to apply the debugging
skills they had honed through years of imperative programming.
The senior partner kept nattering on about QuickCheck properties.
No~progress was made until the first draft of this paper was prepared.
At~that point, we were quickly about to write QuickCheck
specifications for several properties of the Viterbi algorithm:
\begin{itemize}
\item
The total number of \verb+Mat+ and \verb+Ins+ states returned equals
the number of amino acids in the query sequence.
\item
The total number of matches, and deletions returned equals the
number of nodes in the hidden Markov model,
\emph{except} that an initial sequence of
of~\verb+Ins+ states counts as a single node,
due to an odd property of the HMMER file format (namely, the zeroth
node has no delete state and a non-emitting match state which
is always occupied, but a normal insert state).
\item
The sequence returned satisfies the Plan7 invariant, that is, states
\verb+Ins+ and \verb+Del+ are never adjacent.
\end{itemize}
\textbf{These properties need to be followed by more ambitious
properties},
such as proving that nearby sequences are less likely than the
sequence returned by \verb+vee'+.
One practice we recommend for those following in our footsteps is to
write instances of \texttt{Arbitrary} for data types as those data types
are written, when relevant invariants are fresh in memory.

@@

We~modified Matt to use information about sequences as well as structure.
The modification added 2,000 lines of code, and it calls
external sequence aligners that we did not write.
We~thought the modification would take three months, 
but it took most of a~year.
Matt uses such
data structures as mutable oct-trees, vectors, and arrays.
It~uses clever pointer arithmetic.
The mutable data structures were difficult to 
repurpose, and the pointer arithmetic was \emph{too} clever: 
nearly every change resulted in new segfaults.

@@

We~can also make a preliminary comparison of performance.
SMURF can parse a simple protein structure, optimally, in under 10~seconds.
It~can parse some more complex structures in 2--3~minutes.
But when asked to parse the most complex structures, it runs out of
memory and has to be killed, even on a system with 128GB of~RAM.
By~contrast, MRFy can parse any of these structures, approximately,
in 30~seconds.
This performance
meets our goals, and it means that MRFy can be used to scan whole genomes.

Looking beyond our research group to computational biology more
broadly, our experience with other software is better.
Little of it is written in functional languages, 
but much of the software shared by the community is excellent.
MRFy's training component was derived from that of~HMMER,
and
working with the HMMER 
codebase was pleasant;
data structures and their
mutators are well documented. 
There is a 
BioHaskell library, part of which we use,
but it is not nearly as 
complete as BioPython or BioRuby, which are heavily used in the community.
We hope that tools for computational biology in
Haskell continue to mature.

@@

An~anonymous referee spurred us to think
more deeply about laziness.
Josef Svenningsson suggested a correspondence between \mrfy's search
and stream fusion, which led us eventually to the \texttt{Utility} type.
We also thank Lenore Cowen, Kathleen Fisher, Ben Hescott, Brad
Larsen, and Nathan Ricci. 
\fi
